Relational Parametricity for Algebraically Indexed Types

### 2012-06-15 paper/section-introduction.tex
Paragraphs in intro

\section{Introduction}
\label{sec:introduction}

>>>>

\section{Introduction}
\label{sec:introduction}
Types express properties of programs. The statement $\mathit{b}: \tyBool$ tells
us that $\mathit{b}$ has value $\mathtt{true}$ or $\mathtt{false}$;
the statement $\mathit{pos}:\tyReal\tyProduct\tyReal$ says that
$\mathit{pos}$ is a pair whose components are real numbers. More interestingly,
the statement $\mathit{f} : \forall X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that
if we apply $\mathit{f}$ to a list whose elements satisfy some property $\psi$, 
then the result is a list whose elements also satisfy $\psi$.

Unary predicates only go so far, and for polymorphic types
particularly do not express everything that follows from possessing that type. 
Imagine a function $\mathit{oddRev}$ that reverses any list whose first element
is an odd number, otherwise behaving as the identity. It certainly
preserves properties of the elements -- but truly parametric
polymorphism should tell us more.

To obtain stronger properties we must \emph{relate} multiple
executions of the same code. The statement $\mathit{f}:\forall
X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
representation of the list elements supplied to $\mathit{f}$ is
reflected in the elements in the result.  Formally, for any function
$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
essence of \emph{relational parametricity} as formulated by
Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
immediately tells us that $\mathit{oddRev}$ cannot be given a
polymorphic type: simply consider the change-of-representation given
by $\lambda x. x+1$.  But more profoundly it can be used to prove blah
blah (initial algebras, term reps, non-inhabitation, etc.).

The relational interpretation of polymorphism gives us a principle of
\emph{invariance}, namely that generic code is invariant under changes of
representation. It is central to information hiding, abstract data
types, encapsulation and modularity -- at its broadest, software
engineering!  The idea of invariance is also ubiquitous in mathematics
and physics.  The area of a triangle is invariant with respect to
isometries of the Euclidean plane; the determinant of a matrix is
invariant under changes of basis; and Newton's laws are the same in all
inertial frames.  Typically, transformations in mathematics and
physics have interesting structure; for example, translations in the Euclidean plane
form an Abelian group.

Inspired by this connection, we study relational parametricity for 
types that are \emph{indexed} by attributes that themselves have algebraic structure.
For example, in geometry, points in 2-space can be indexed by a co-ordinate frame;
in physical modelling, quantities are indexed by units-of-measure;
in information-flow security, computations are indexed by principals. 
Types that are polymorphic over such indices induce invariance properties. 
A function that computes the area of a triangle might have simple type
\[
\mathit{area} : \tyPrimNm{vec}\tyArr\tyPrimNm{vec}\tyArr\tyPrimNm{vec}\tyArr\tyReal
\]
but by assigning a polymorphic type
\[
\mathrm{area} : \forall i\mathord:\SynSE{2}.
  \tyPrim{vec}{i} \to \tyPrim{vec}{i} \to \tyPrim{vec}{i} \to \tyReal \\
\]
we can deduce that it is invariant under direct isometries (rigid body motions) as 
represented by the index $i$ on the type of vectors.
\todo{polymorphism imposes abstraction barrier; talk about co-ordinate free geometry}
